---
title: "250919-Application of Glasso on INT-Transformed Zero-Inflated Data"
author: "Jingxin Wang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Last Time

1. Pearson partial correlation is more sensitive to transformation than Spearman and Kendall
2. Normality satisfied after INT transformation.
3. We can experiment Glasso on INT data, and compare the result with network given by Spearman and Kendall
4. Multiple test correction for p-value based edge selection should be tested.


## 1. Today

1. Application of Glasso on zero-inflated data before and after INT transformation
  - Compare the networks before and after transformations
2. Benchmark with Hypothesis Testing
  - Assumptions check for Benjamin-Hotchberg and Benjamin-Yekutielli Test, do they work for correlation of the current type?
  
## 2. Loading Data and Packages

```{r message=FALSE}
## Load packages
library(HurdleNormal)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(igraph)
library(tidygraph)
library(ggraph)

## Load customized functions
source("../R-scripts/INT.R")

## Load and Sample Methylation Data
X.T <-  readRDS(file = "../../Data/Common_pan_cancer_hyper_bins_adjusted_and_normalized_cnt_in_SE.RDS")
set.seed(226)
sampDMR <- sample(rownames(X.T), 10) ## Sample only 10 DMRs for speed sake
original.X <- t(X.T[sampDMR, ])

## INT Transformation
INT.X <- INT(original.X, 0.5)
```



## 3. Glasso on Zero-Inflated the Data, Before and After INT

```{r}


```


## 4. Benjamin-Hotchberg and Benjamin-Yekutielli Test for Edge Selection


