---
title: "251007-Application of Glasso on INT-Transformed Zero-Inflated Data"
author: "Jingxin Wang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 0. Last Time

1. Pearson partial correlation is more sensitive to transformation than Spearman and Kendall
2. Normality satisfied after INT transformation.
3. We can experiment Glasso on INT data, and compare the result with network given by Spearman and Kendall
4. Multiple test correction for p-value based edge selection should be tested.


## 1. Today

1. Application of Glasso on zero-inflated data before and after INT transformation
  - Compare the networks before and after transformations
2. Can we compare results of different transformation approaches with same Glasso rho?  
  - How to put methylation before and after transformation into the same scale?
  - As rho increases, do paths of precision matrix elements change much before and after INT?  
    * BIC trace curves: healthy vs cancers
    * Path of the precision matrix component
  - Can we center every column of the matrix and scale it by the same factor before Glasso?
```{r}
## Load packages
library(tidyverse)
library(ggplot2)
library(cowplot)
library(igraph)
library(tidygraph)
library(ggraph)
library(ppcor)
library(iDINGO)
library(HurdleNormal)

## Define Global functions
## Extract upper triangle part of matrices
takeUpperTriangle <- function(Mat) {
    Vec <- Mat[upper.tri(Mat)]
    as.vector(Vec)
}

## Compare the partial correlations before and after transformations for all types of correlations
compareParScatter <- function(refPar, expPar, title) {
  ggplot(data = data.frame("Before" = refPar, "After" = expPar)) + 
      geom_point(aes(x = Before, y = After)) + 
      coord_cartesian(xlim = c(-0.4,0.4), ylim = c(-0.4,0.4)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
      labs(title = title)
}


## Inverse Component Trajectory and eBIC path
precMatPath <- function(X, rho.Vec, transName) {
  p <- ncol(X); n <- nrow(X)
  
  upperTriMat <- matrix(nrow = 0, ncol = p*(p-1)/2)
  eBIC.vec <- c()
  for (rho in rho.Vec) {
    glassoMat <- glassoPreMat(original.X, rho = rho)
    upperTriVec <- takeUpperTriangle(glassoMat$wi)
    upperTriMat <- rbind(upperTriMat, upperTriVec)
    eBIC <- extendedBIC(0.5, glassoMat$wi, cov(original.X), n)
    eBIC.vec <- c(eBIC.vec, eBIC)
  }
  
  par(mfrow = c(1,2))
  matplot(x = rho.Vec, y = upperTriMat, type = "l", lty = 1,
          xlab = "rho", ylab = "Precision Matrix Coef",
          ylim = c(-0.3, 0.3), main = paste("Precision matrix element path of", transName))
  
  plot(x = rho.Vec, y = eBIC.vec, type = "l", lty = 1,
        xlab = "rho", ylab = "Extended BIC",
        main = paste("Extended BIC of", transName))
}
```

## 2. Loading Data and Packages

```{r message=FALSE}
## Load customized functions
source("../R-scripts/INT.R")

## Load and Sample Methylation Data
X.T <-  readRDS(file = "../../Data/Common_pan_cancer_hyper_bins_adjusted_and_normalized_cnt_in_SE.RDS")
set.seed(226)
sampDMR <- sample(rownames(X.T), 10) ## Sample only 10 DMRs for speed sake
original.X <- t(X.T[sampDMR, ])

## Load Cancer Types
sampleInfo <- read.csv("../../Data/Common_pan_cancer_hyper_bins_adjusted_cnt_in_SE_samples_info.csv")
cancerTypes <- sampleInfo$cancer_type

## INT Transformation
INT.X <- INT(original.X, 0.25)
```



## 3. Glasso on Zero-Inflated the Data, Before and After INT, All Samples
### 3.1 Network Comparison

```{r}
source("../R-scripts/graphMaker.R")
library(glasso)

# original.X <- scale(original.X, center = T, scale = F)/50 ## Centering and Scaling by a factor of 50

## Partial Pearson Correlation before any transformations
ParCorrNone <- pcor(original.X, method = "pearson")

## Partial Pearson Correlation after INT transformations
ParCorrINT <- pcor(INT.X, method = "pearson")

## Glasso before INT
glassoNone <- glassoPreMat(original.X, rho = 0.25)
ParCorrGlassoNone <- parCorr_from_precision(glassoNone$wi, 0.25, nrow(original.X))

## Glasso after IVT
glassoINT <- glassoPreMat(INT.X, rho = 0.25)
ParCorrGlassoINT <- parCorr_from_precision(glassoINT$wi, 0.25, nrow(INT.X))

## Graphical Comparison
ParCorrNoneGraph <- plot_parcorr_graph(ParCorrNone$estimate, NA, "All", "Before INT")
ParCorrINTGraph <- plot_parcorr_graph(ParCorrINT$estimate, NA, "All", "After INT")
ParCorrNoneGlassoGraph <- plot_parcorr_graph(ParCorrGlassoNone, 0.25, "All", "Before INT Glasso")
ParCorrINTGlassoGraph <- plot_parcorr_graph(ParCorrGlassoINT, 0.25, "All", "After INT Glasso")

plot_grid(plotlist = list(ParCorrNoneGraph, ParCorrINTGraph,
                          ParCorrNoneGlassoGraph, ParCorrINTGlassoGraph), 
          nrow = 2, ncol = 2)
ggsave('../../Images/GraphVis/AllSampleINTGlasso.png')
```

### 3.2 Does INT significantly alter precision matrix components?

Next, we will use scatter plot to compare the strength of upper triangular components of precision matrices before and after INT transformation.
```{r}
## Precision Matrix without penalty
condcenter.X <- conditionalCenter(original.X)
originalPrecMat <- solve(cov(condcenter.X))
intPrecMat <-solve(cov(INT.X))

originalCompo <- takeUpperTriangle(originalPrecMat)
intCompo <- takeUpperTriangle(intPrecMat)

## Precision Matrix with penalty
originalGlasso <- takeUpperTriangle(glassoNone$wi)
intGlasso <- takeUpperTriangle(glassoINT$wi)

## Plotting
compareParScatter(originalCompo, intCompo, "Precision Matrix")
compareParScatter(originalGlasso, intGlasso, "Glasso Matrix")

## Which component deviated from the original precision the most?
cat("For unpenalized precision matrix --------------------\n")
cat("The ", which.max(abs(originalCompo - intCompo)), "th component shows the biggest error of ", max(abs(originalCompo - intCompo)), "\n")
cat("As compared maximum absolute value of all components before transformation:", max(abs(originalCompo)), "\n")
cat("For glasso precision matrix ----------------------\n")
cat("The ", which.max(abs(originalGlasso - intGlasso)), "th component shows the biggest error of ", max(abs(originalGlasso - intGlasso)), "\n")
cat("As compared maximum absolute value of all components before transformation:", max(abs(originalGlasso)), "\n")

abs(originalPrecMat - intPrecMat) == max(abs(originalCompo - intCompo))

cat("Before INT, the 19th correlation is:", cor(original.X[, "523070"], original.X[, "8978709"]))
cat("After INT, the 19th correlation is:", cor(INT.X[, "523070"], INT.X[, "8978709"]))
```

```{r}
precMatPath(original.X, seq(0.1, 5, 0.1), "Original")
precMatPath(INT.X, seq(0.1, 5, 0.1), "INT")
```

**Main Takeaway**:
1. INT transformation can lead to big differences in partial correlations from the original data. INT was not used in rhe right way or it simply distorted true partial correlation. 
2. Convergence of all upper triangular components occur at rho = 2

**Observations**:
1. For unpenalized Pcorr graphs, egde sets look similar before and after INT, but the color lightens after INT.
2. For glasso graphs, not only color lightens, but also removal of heavily weighted old edges and replacement by new and lighter weighted edges.
3. The error in inverse covariance matrix components before and after transformation can be very large relative to the size of the component before transformation.
4. The component giving the biggest error in the regular precision matrix also gives the biggest error in the glasso matrix; the component locate at an off-diagonal entry.
5. There is very difference in the evolution path between original and INT-transformed glasso components.
6. As rho increases, eBIC increases without any sign of slowing the increase. This could imply a worsening representation of the original partial correlation structure as the model grow sparser.

### 3.3 Does Conditional Centering significantly alter precision matrix?

```{r}
## Precision Matrix without penalty
condcenter.X <- conditionalCenter(original.X)
condCenterPrecMat <-solve(cov(condcenter.X))
condCenterCompo <- takeUpperTriangle(condCenterPrecMat)

## Plotting
compareParScatter(originalCompo, condCenterCompo, "Precision Matrix")

## Which component deviated from the original precision the most?
cat("For unpenalized precision matrix --------------------\n")
cat("The ", which.max(abs(originalCompo - condCenterCompo)), "th component shows the biggest error of ", max(abs(originalCompo - condCenterCompo)), "\n")
cat("As compared maximum absolute value of all components before transformation:", max(abs(originalCompo)), "\n")
abs(originalPrecMat - condCenterPrecMat) == max(abs(originalCompo - condCenterPrecMat))

```
Conditional centering of non-zero values did not alter precision matrix components as much as INT.

## 4. HurdleNormal with Different the Transformation Methods: Original, INT+Repositioning Hurdle, Conditional Centering
```{r}
## INT + Repositioning the Hurdle at 0
INT.X.reposit <- apply(INT.X, MARGIN = 2, FUN = function(x) x - min(x))
hist(INT.X.reposit[,1], main = "Marginal distribution after moving the hurdle to 0", xlab = "Y1")
```
Conditional centering will substract the mean of non-zero values from original value.

```{r}
condcenter.X <- conditionalCenter(original.X)
plot(condcenter.X[,1], condcenter.X[,3], main = "Joint Distribution of Y1, Y2 after condition centering", xlab = "Y1", ylab = "Y2")
```


```{r}
hurdleComparison <- function(cancerType) {
  hurdleFit.original <- fitHurdle(original.X[cancerTypes == cancerType, ], nlambda = 10)
  hurdleFit.INT <- fitHurdle(INT.X.reposit[cancerTypes == cancerType, ], nlambda = 10)
  hurdleFit.condcenter <- fitHurdle(condcenter.X[cancerTypes == cancerType, ], nlambda = 10)
  
  ## Compare path of extended BIC
  candidateModels <- list("Original" = hurdleFit.original, 
                          "INT+Reposition" = hurdleFit.INT, 
                          "Conditional Centering" = hurdleFit.condcenter)
  bicPlotList <- list()
  for (i in 1:3) {
    bicPlotList <- append(bicPlotList, 
                          ggplot(data = candidateModels[[i]]$BIC_etc)+
                            geom_line(aes(x = trueEdges, y = BIC))+
                            labs(title = names(candidateModels)[i]))
  }
  
  plot_grid(plotlist = bicPlotList, nrow = 1) %>% print()

  ## Visualize the coef path
  coefPathList <- list()
  par(mfrow = c(1,3))
  for (i in 1:3) {
    coefPath <- lapply(candidateModels[[i]]$adjMat, FUN = takeUpperTriangle) %>%
      reduce(.f = rbind)
    rhoPath <- candidateModels[[i]]$lambda
    
    matplot(x = rhoPath, y = coefPath, type = "l", lwd = 2,
            main = paste("Coef Paths:", names(candidateModels)[i]),
            xlab = "rho", ylab = "Off-diagonal Entries")
  }
  
  ## Network Comparison
  positList <- c(1,3,8,10)
  
  networkPlotList <- list()
  for (i in positList) {
    originalGraph <- plot_parcorr_graph(candidateModels[[1]]$adjMat[[i]] %>% as.matrix(), 
                                          candidateModels[[1]]$lambda[i] %>% round(2),
                                          "Healthy", "Original")
    intGraph <- plot_parcorr_graph(candidateModels[[2]]$adjMat[[i]] %>% as.matrix(), 
                                        candidateModels[[2]]$lambda[i] %>% round(2),
                                        "Healthy", "INT")
    condCenterGraph <- plot_parcorr_graph(candidateModels[[3]]$adjMat[[i]] %>% as.matrix(), 
                                        candidateModels[[3]]$lambda[i] %>% round(2),
                                        "Healthy", "Conditional Centering")
      
    networkPlotList <- append(networkPlotList, 
                          plot_grid(plotlist = list(originalGraph, intGraph, condCenterGraph),
                                    nrow = 1))
  }
  
  plot_grid(plotlist = networkPlotList[1:2], ncol = 1, nrow = 4) %>% print()
  plot_grid(plotlist = networkPlotList[1:2], ncol = 1, nrow = 4) %>% print()
  gc()
  return(0)
}

```

```{r message=F}
hurdleComparison("Normal")
hurdleComparison("Blood Cancer")
```



## 5. Do edge sets differ a lot between healthy and cancers?

```{r}
allTypes <- unique(cancerTypes)
allTypes <- allTypes[allTypes != "Normal"]

## Preparing healthy normal data
healthy.original <- original.X[cancerTypes == "Normal", ]
healthy.INT <-  INT(healthy.original, 0.25)

## Exact precision matrix
precMat.healthyOri <- solve(cov(healthy.original))
precMat.healthyINT <- solve(cov(healthy.INT))


## Take Uppper Triangular compoents
precCompo.healthyOri <- takeUpperTriangle(precMat.healthyOri)
precCompo.healthyINT <- takeUpperTriangle(precMat.healthyINT)

```

```{r}
for (k in allTypes) {
  ## Prepare precision matrices for subset of data based on cancer types.
  X_k.original <- original.X[cancerTypes == k, ]
  X_k.INT <- INT(X_k.original, 0.25)
  
  precMat.original <- solve(cov(X_k.original))
  precMat.INT <- solve(cov(X_k.INT))
  
  ## Scatter Plot Comparison
  precCompo.original <- takeUpperTriangle(precMat.original)
  precCompo.INT <- takeUpperTriangle(precMat.INT)
  
  
  A <- ggplot(data = data.frame("Healthy" = precCompo.healthyOri, "Cancer" = precCompo.original)) + 
      geom_point(aes(x = Healthy, y = Cancer)) + 
      # coord_cartesian(xlim = c(-0.4,0.4), ylim = c(-0.4,0.4)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
      labs(title = paste("Healthy vs ", k, "Original"))
  
  B <- ggplot(data = data.frame("Healthy" = precCompo.healthyINT, "Cancer" = precCompo.INT)) + 
      geom_point(aes(x = Healthy, y = Cancer)) + 
      # coord_cartesian(xlim = c(-0.4,0.4), ylim = c(-0.4,0.4)) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
      labs(title = paste("Healthy vs ", k, ", INT"))
  
  plot_grid(plotlist = list(A, B), nrow = 1, ncol = 2) %>% print()
}
```

**Observations**:
1. Edge strengths of different cancers very different those of healthy normals.
2. Precision matrix components of Bladder cancer tend to concentrate around 0, with 3 outliers outside [-2, 2] band.
3. Linear relationship in regular precision matrix entries were observed between healthy sample and colorectal, pancreatic, and renal cancer sample
